---
title: "Wrangling Temporal Data"
subtitle: "Fundamentals of Data Science"
author: "MTH-391A | Spring 2025 | University of Portland"
date: "March 12, 2025"
output:
  slidy_presentation:
    font_adjustment: +5
    footer: "| MTH-391A Spring 2025 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
bibliography: ../../references.bib
csl: ../../mbe.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(gridExtra)
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- 
- **Activity: **
::::

:::: {.column width=15%}
::::

## What is Temporal Data?

**Temporal data** refers to data that is associated with time, meaning it captures changes, trends, or patterns over a period.

**Key Characteristics:**

* *Time-Stamped:* Each data point is linked to a specific time (e.g., a date or timestamp).
* *Sequential Order:* The order of data points matters, as it reflects change over time.
* *Trends & Patterns:* Useful for analyzing progression, seasonality, or recurring behaviors.

## Case Study I

**New York City Flights**

* The `nycflights13` package: On-time data for all flights that departed NYC (i.e. JFK, LGA or EWR) in 2013
* The `nycflights23` package: Similar to `nycflights13` but in 2023

**Load Packages**

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(nycflights13)
library(nycflights23)
```

**Data Frames**

```{r echo=TRUE, message=FALSE, warning=FALSE}
# rename the data frames since both packages have identical data frame names
flights13 <- nycflights13::flights
flights23 <- nycflights23::flights
```

## Time as a Dimension/Variable



## Time is Relative



## Time Formats



## Wrangling Time Variables



## Organizing Time for Analysis



## Activity: 

1. Log-in to Posit Cloud and open the R Studio assignment *MA14: *.
2. Make sure you are in the current working directory. Rename the `.Rmd` file by replacing `[name]` with your name using the format `[First name][Last initial]`. Then, open the `.Rmd` file.
3. Change the author in the YAML header.
4. Read the provided instructions.
5. Answer all exercise problems on the designated sections.
