---
title: "Working with Date/Time Variables"
subtitle: "Fundamentals of Data Science"
author: "[name]"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

* You must submit your mini-assignment individually by end-of-class or end-of-day. Your name must exist in your mini-assignment and the names of your collaborators.
* Mini-assignments are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on mini-assignments, and no opportunities for revisions and make-up.

# Instructions

* For the exercises, please provide your answers in full sentences.
* Write your answers by replacing the text "[Write your answer here]".
* Please read the problems fully and carefully before answering.
* When finished, knit this `.Rmd` to `.html`.
* Submit this `.Rmd` and the recently knitted `.html`.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(nycflights13)
library(nycflights23)
```

# Collaborators

::: {style="color: blue;"}
[Write your answer here]
:::

# Exercises

## Problem 1

What happens if you parse a string that contains invalid dates? Explain why.

```{r}
ymd(c("2010-10-10", "bananas"))
```

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 2

For each of the following dates and times, show how you would process it using the date-time specification functions to convert it into date/time format.

```{r}
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014
t1 <- "1705"
t2 <- "11:15:10.12 PM"
```

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 3

This problem uses the following merged data from the `nycflights13` and `nycflights23` packages.

```{r echo=TRUE, message=FALSE, warning=FALSE}
# bind the two data frames
flights <- nycflights13::flights %>% 
  rbind(nycflights23::flights)
```

Using the `flights` data frame, create a subset of flights that departed between November 12 and December 24 for both years 2013 and 2023.

::: {style="color: blue;"}
[Write your answer here]
:::
