---
title: "Combine Data Sets"
subtitle: "Fundamentals of Data Science"
author: "[First name][Last initial]"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

* You must submit your worksheet individually by end-of-class or end-of-day. Your name must exist in your worksheet and the names of your collaborators.
* Worksheets are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on worksheets, and no opportunities for revisions and make-up.

# Instructions

* For the exercises, please provide your answers in full sentences.
* Write your answers by replacing the text "[Write your answer here]".
* Please read the problems fully and carefully before answering.
* When finished, knit this `.Rmd` to `.html`.
* Submit this `.Rmd` and the recently knitted `.html`.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(Lahman)
```

# Collaborators

::: {style="color: blue;"}
[Write your answer here]
:::

# Exercises

## Problem 1

Consider the following data frames with information about U.S. states from 1977.

```{r}
statenames <- tibble(names = state.name, twoletter = state.abb)
glimpse(statenames)
```

```{r}
statedata <- tibble(
  names = state.name, 
  income = state.x77[, 2], 
  illiteracy = state.x77[, 3]
  )
glimpse(statedata)
```

Join the two data frames by matching the states.

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 2

Use the `Batting`, `Pitching`, and `People` tables in the `Lahman` package to answer the following questions. Here, you may need to use verbs from the `dplyr` package along with `inner_join()`, `left_join()`, or `right`_join()`.

a. Name every player in baseball history who has accumulated at least 300 home runs (`HR`) and at least 300 stolen bases (`SB`). You can find the first and last name of the player in the `People` data frame. Join this to your result along with the total home runs and total bases stolen for each of these elite players.

b. Similarly, name every pitcher in baseball history who has accumulated at least 300 wins (`W`) and at least 3,000 strikeouts (`SO`).

c. Identify the name and year of every player who has hit at least 50 home runs in a single season. Which player had the lowest batting average in that season?

::: {style="color: blue;"}
[Write your answer here]
:::
