---
title: "Visualize High-Dimensional Data"
subtitle: "Fundamentals of Data Science"
author: "[name]"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

* You must submit your mini-assignment individually by end-of-class or end-of-day. Your name must exist in your mini-assignment and the names of your collaborators.
* Mini-assignments are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on mini-assignments, and no opportunities for revisions and make-up.

# Instructions

* For the exercises, please provide your answers in full sentences.
* Write your answers by replacing the text "[Write your answer here]".
* Please read the problems fully and carefully before answering.
* When finished, knit this `.Rmd` to `.html`.
* Submit this `.Rmd` and the recently knitted `.html`.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(palmerpenguins)
```

# Collaborators

::: {style="color: blue;"}
[Write your answer here]
:::

# Exercises

## Problem 1

Apply Principal Component Analysis (PCA) to compute the components of the four numerical variables `bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, and `body_mass_g` in the `penguins` data set.

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 2

Plot the explained variance and the cumulative explained variance.

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 3

What is the total variance explained for the top two principal components? What does it mean? Explain your reasoning.

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 4

Visualize the two principal components computed in Problem (1) using three separate scatter plots. In each plot, the points should be colored based on the following categorical variables:

* `species`
* `island`
* `sex`

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 5

Explain your observations in each plot in Problem (4), and describe their differences.

::: {style="color: blue;"}
[Write your answer here]
:::
