---
title: "Working with Geometry Variables"
subtitle: "Fundamentals of Data Science"
author: "[name]"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

* You must submit your mini-assignment individually by end-of-class or end-of-day. Your name must exist in your mini-assignment and the names of your collaborators.
* Mini-assignments are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on mini-assignments, and no opportunities for revisions and make-up.

# Instructions

* For the exercises, please provide your answers in full sentences.
* Write your answers by replacing the text "[Write your answer here]".
* Please read the problems fully and carefully before answering.
* When finished, knit this `.Rmd` to `.html`.
* Submit this `.Rmd` and the recently knitted `.html`.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(ggthemes)
library(usdata)

# mapping related packages
library(scales)
library(usmap)
library(urbnmapr)
library(sf)
```

# Collaborators

::: {style="color: blue;"}
[Write your answer here]
:::

# Exercises

## Problem 1

Below, we demonstrate how to create maps using the `usmap` package. This package is specifically for making US maps for national, state, and county levels. For the R code chunk below, examine the R code and solve the following:

a. Choose a US state to map.
b. Choose a color sequential scheme from [ColorBrewer 2.0](https://colorbrewer2.org/#type=sequential&scheme=PuRd&n=6){target="_blank"}.
    
Make sure to adjust your figure widths and heights to avoid distortions or overlapping labels. Once you draw your map, describe your observations.

```{r}
# New variable where the 2022 population is converted into an 
#   ordinal categorical variable with n levels.
#   Here we are using the countypop dataframe
n_levels <- 6
pop_level <- cut_number(log10(countypop$pop_2022),n_levels)
county_new <- countypop %>% 
  mutate(population = pop_level)

# Map a state with county borders

#### Start: (1) Choose a US State here
state_full <- "California"
state_abrv <- "CA"
#### End: (1) Choose a US State here

plot_usmap(data = county_new, 
           values = "population", 
           include = c(state_abrv)) + # uses the state abbv
  
#### Start: (2) Change color scheme here 
  # number of colors must match number of levels - these are HEX color codes
  scale_fill_manual(values = 
                       c('#ffffcc','#d9f0a3','#addd8e',
                         '#78c679','#31a354','#006837')) + 
#### Start: (2) Change color scheme here
  
  guides(fill=guide_legend(nrow=2,byrow=TRUE)) + 
  labs(fill = "log10(population)",
       title = paste("2022 Population County-Level Map of",state_full)) +
  theme(legend.position = "bottom")
```

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 2

Below, we demonstrate another way to map a US state using the `sf` package, which it can extend shapefiles into dataframe-like objects. Here, we are using the shapefiles provided by the [`urbnmapr`](https://urbaninstitute.github.io/urbnmapr/index.html){target="_blank"} package, which is from the [Urban Institute](https://www.urban.org/){target="_blank"}. For the R code chunk below, examine the R code and solve the following:

a. Choose a US state to map other than what you chose in problem 1.
b. Use the `county` data set from the `usdata` package where you use the `unemployment_rate` variable and convert it into an ordinal categorical variable with 6 levels.
c. Combine the shapefile dataframe and part (b)s dataframe where you match the county and the state names.
d. Plot the county-level state map where you color each county proportional to the ordinal categorical variable done in part (b). Choose a color sequential scheme from [ColorBrewer 2.0](https://colorbrewer2.org/#type=sequential&scheme=PuRd&n=6){target="_blank"}.
    
Make sure to adjust your figure widths and heights to avoid distortions or overlapping labels. Once you draw your map, describe your observations.

```{r message=FALSE, warning=FALSE}
#### Start: (1) Choose a US State here
state_full <- "California"
#### End: (1) Choose a US State here

# Load the shapefile object of US counties and filter only the state of choice
#   using the urbnmapr package
counties_sf <- get_urbn_map("counties", sf = TRUE)
counties_sf_ca <- counties_sf %>% filter(state_name == state_full)
```

```{r}
#### Start: (2) Wrangle the county data set here
county_sub <- county %>%
  select(name,state,unemployment_rate)
#### End: (2) Wrangle the county data set here

#### Start: (3) Update the ggplot pipeline here
ggplot(data = counties_sf_ca, aes()) + 
  geom_sf() +
  labs(x = "latitude",
       y = "longitude",
#### End: (3) Update the ggplot pipeline here
       title = paste("Unemployment Rate County-Level Map of",state_full))
```

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 3

Below, we demonstrate another way to map a US state, were we use an actual shapefile. For the R code chunk below, examine the R code and solve the following:

a. Use the `full_join()` function to combine the two shapefiles where you match the state and county codes.
b. Choose a state and map it while coloring the counties using the `poverty` variable from the `county` data set.
    
Make sure to adjust your figure widths and heights to avoid distortions or overlapping labels. Once you draw your map, describe your observations.

```{r}
# US Census Bureau 2018 county-level shapefile files (Downloaded)
# source: 
#   https://www.census.gov/geographies/mapping-files/
#     time-series/geo/carto-boundary-file.html
# Downloaded files are included in the zip for for this mini assignment

# load the shapefiles
state_shp <- read_sf("cb_2018_us_state_20m/cb_2018_us_state_20m.shp")
county_shp <- read_sf("cb_2018_us_county_20m/cb_2018_us_county_20m.shp")
```

::: {style="color: blue;"}
[Write your answer here]
:::
