---
title: "Visualizing Variables and Locations on A Map"
subtitle: "Fundamentals of Data Science"
author: "[name]"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

* You must submit your mini-assignment individually by end-of-class or end-of-day. Your name must exist in your mini-assignment and the names of your collaborators.
* Mini-assignments are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on mini-assignments, and no opportunities for revisions and make-up.

# Instructions

* For the exercises, please provide your answers in full sentences.
* Write your answers by replacing the text "[Write your answer here]".
* Please read the problems fully and carefully before answering.
* When finished, knit this `.Rmd` to `.html`.
* Submit this `.Rmd` and the recently knitted `.html`.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(ggthemes)
library(usdata)

# mapping related packages
library(maps)
library(usmap)
library(urbnmapr)
library(sf)
```

# Collaborators

::: {style="color: blue;"}
[Write your answer here]
:::

# Exercises

## Problem 1

we are using the shapefiles provided by the [`urbnmapr`](https://urbaninstitute.github.io/urbnmapr/index.html){target="_blank"} package, which is from the [Urban Institute](https://www.urban.org/){target="_blank"}. For the R code block below, examine the R code and solve the following:

a. Choose a US state to map other than the example.
b. Use the `county` data set from the `usdata` package where you use the `pop_change` variable and convert it into an ordinal categorical variable with 6 levels.
c. Combine the shapefile dataframe and part (b)s dataframe where you match the county and the state names.
d. Plot the county-level state map where you color each county proportional to the ordinal categorical variable done in part (b). Choose a color diverging scheme from [ColorBrewer 2.0](https://colorbrewer2.org/#type=sequential&scheme=PuRd&n=6){target="_blank"}.
e. Use the `us.cites` data set from the `maps` package to put point onto your map.
    
Make sure to adjust your figure widths and heights to avoid distortions or overlapping labels.
Once you draw your map, describe your observations.

```{r message=FALSE}
#### Start: (1) Choose a US State here
state_full <- "California"
state_abv <- "CA"
#### End: (1) Choose a US State here

# Load the shapefile object of US counties and filter only the state of choice
#   using the urbnmapr package
counties_sf <- get_urbn_map("counties", sf = TRUE)
counties_sf_ca <- counties_sf %>% filter(state_name == state_full)
```

```{r state_county_pop_map_2, fig.align='center', out.width="100%", fig.height=7}
#### Start: (2) Wrangle the county data set here
# subset of the county data set
county_sub <- county %>%
  select(name,state,pop_change)
# cities data set
crs_use <- "+proj=laea +lat_0=30 +lon_0=-95" # Albers projection with curvature
us_cities <- us.cities %>%
  mutate(name_n = str_remove_all(name, state_abv)) %>%
  mutate(name_n = str_remove_all(name_n, " ")) %>%
  filter(country.etc == state_abv)
d_points <-  us_cities %>% 
  st_as_sf(coords = c("long", "lat"), crs = 4326) %>% 
  st_transform(crs = crs_use)
#### End: (2) Wrangle the county data set here

#### Start: (3) Update the ggplot pipeline here
ggplot(data = counties_sf_ca, aes()) + 
  geom_sf() + 
  labs(title = paste("Population Change County-Level Map of",state_full)) +
#### End: (3) Update the ggplot pipeline here
theme(legend.position="bottom",
        panel.background=element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank())
```

::: {style="color: blue;"}
[Write your answer here]
:::
