---
title: "Visualizing and Analyzing Temporal Data"
subtitle: "Fundamentals of Data Science"
author: "MTH-391A | Spring 2025 | University of Portland"
date: "March 12, 2025"
output:
  slidy_presentation:
    font_adjustment: +5
    footer: "| MTH-391A Spring 2025 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
bibliography: ../../references.bib
csl: ../../mbe.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(gridExtra)
set.seed(42)
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- **Know how to visualize time-series in multiple ways**
- **Develop an understanding on how different time-series visualizations affect interpretation.**
- **Activity: Visualize and Interpret Time-Series**
::::

:::: {.column width=15%}
::::

## Previously... (1/2)

**Date/Time Ambiguity and Formats**

:::: {.column width=49%}
**Example Dates**

```{r echo=TRUE}
# year-month-day
ymd("2017-01-31")

# month-day-year
mdy("January 31st, 2017")

# day-month-year
dmy("31-Jan-2017")
```
::::

:::: {.column width=50%}
**Converting String Dates into date/time format**

* Identify the order in which year, month, and day appear in your dates
* Arrange “y”, “m”, and “d” in the same order
* That gives you the name of the function that will parse your date into the [ISO8601](https://www.iso.org/iso-8601-date-and-time-format.html){target="_blank"}, an international standard for writing date/time.
::::

## Previously... (2/2)

**Time Granularity and Time-Zones**

* **Granularity Matters:** The level of detail or precision used to represent time in a given context
    - Seconds, minutes, hours, days, weeks, months, years, decades, etc.
    - Semesters (fall, spring)
    - Seasons (winter, spring, summer, fall)
    - TV Show seasons or movie sequels
    - Election cycles
* **Time Zones Shift Meaning:** Be mindful of time-zones
    - A "morning sale" in New York City, USA isn’t morning in Manila, Philippines

## Case Study I

**New York City Flights**

* The `nycflights13` package: On-time data for all flights that departed NYC (i.e. JFK, LGA or EWR) in 2013
* The `nycflights23` package: Similar to `nycflights13` but in 2023

:::: {.column width=49%}
**Load Packages**

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(nycflights13)
library(nycflights23)
```
::::

:::: {.column width=50%}
**Data Frames**

```{r echo=TRUE, message=FALSE, warning=FALSE}
# bind the two data frames
flights <- nycflights13::flights %>% 
  rbind(nycflights23::flights)

# show size
dim(flights)
```
::::

## Variables of Interest

Each row in the `flights` tibble is a unique flight from NYC.

* `date_time` | date and time of departure, which can be decomposed into parts (`year`, `month`, `day`, `hour`)
* `carrier` | type of airline
* `origin` | origin airport in New York (JFK, LGA or EWR)
* `dest` | destination airport (e.g. ATL, LAX, PDX)
* `dep_delay` | Departure delays, in minutes. Negative times represent early departures/arrivals.

```{r echo=FALSE}
# create subset
flights_sub <- flights |> 
  mutate(date_time = make_datetime(year,month,day,hour)) |> 
  select(date_time,carrier,origin,dest,dep_delay,year,month,day,hour)
```

```{r echo=TRUE}
# view random sample
flights_sub |> 
  sample_n(3)
```

## How to Get Started on Exploring Temporal Data

The main **goal of time-series data** is to analyze how a variable or set of variables evolve over time. This typically involves identifying patterns and trends.

**How many flights occurred in 2013 and 2023?**

```{r echo=FALSE, fig.width=4,fig.height=3, fig.align='center'}
ggplot(flights_sub,aes(x=factor(year))) + 
  geom_bar() + 
  xlab("year") + 
  ylab("count")
```

## Increase Granularity

**What is the number of fights per month?**

```{r echo=FALSE, fig.width=6,fig.height=2, fig.align='center', out.width='100%'}
ggplot(flights,aes(x=factor(month),fill=factor(year))) + 
  geom_bar(position="dodge") + 
  xlab("month") + 
  ylab("count") + 
  guides(fill=guide_legend(title="year"))
```

## Incorporate Other Variables

**What is median departure delay per month coming out of each NYC airport?**

```{r echo=FALSE, fig.width=6,fig.height=3, fig.align='center', out.width='100%'}
flight_median_delay <- flights %>% 
  drop_na() %>%
  group_by(year,month,origin) %>% 
  summarise(median_delay = median(dep_delay),
            .groups="drop")

ggplot(flight_median_delay,aes(x=month,y=median_delay,color=origin)) + 
  geom_point() + 
  geom_line() + 
  xlab("month") + 
  ylab("median departure delay") + 
  facet_wrap(~year,nrow=3,scales="fixed") +
  guides(color=guide_legend(title="origin")) + 
  scale_x_continuous(breaks=seq(1,12,1))
```

## More Precision

**What is the median daily delay during the summer months?**

```{r echo=FALSE, fig.width=8,fig.height=4, fig.align='center', out.width='100%'}
flight_median_delay_byday <- flights %>% 
  drop_na() %>%
  group_by(year,month,day,origin) %>% 
  summarise(median_delay = median(dep_delay),
            .groups="drop")

ggplot(flight_median_delay_byday %>% filter(month %in% c(6,7,8,9),origin=="JFK"),aes(x=factor(day),y=median_delay,color=factor(month),group=factor(month))) + 
  geom_point() + 
  geom_line() + 
  scale_color_brewer(palette = "RdYlGn") + 
  xlab("day of the month") + 
  ylab("median delay") + 
  facet_wrap(~year,nrow=3,scales="fixed") + 
  guides(color=guide_legend(title="origin")) 
```

## A Stragety on Visualizing Time-Series Data

**Effective Methods for Analyzing Trends Over Time**

1. **Define the Objective**
  
  - What insights are you looking for?
  - Are you analyzing trends, seasonality, anomalies, or forecasting?
  
2. **Choose the Right Visualization Type**

  - Line graphs: Best for continuous trends over time
  - Bar Plots: Good for comparing discrete time intervals
  - Boxplots: Helps visualizing distributions for comparing discrete time intervals
  
3. **Data Preprocessing**

  - Handle missing values.
  - Aggregate data to appropriate time intervals (hourly, daily, monthly).
  - Normalize or scale data if necessary.
  
4. **Enhance Readability**
  
  - Use consistent time intervals on the x-axis.
  - Add annotations for significant events.
  - Utilize color coding for multiple time series.

## Activity: Visualize Time-Series

1. Log-in to Posit Cloud and open the R Studio assignment *MA15: Visualize Time-Series*.
2. Make sure you are in the current working directory. Rename the `.Rmd` file by replacing `[name]` with your name using the format `[First name][Last initial]`. Then, open the `.Rmd` file.
3. Change the author in the YAML header.
4. Read the provided instructions.
5. Answer all exercise problems on the designated sections.
