---
title: "Visualize Time-Series"
subtitle: "Fundamentals of Data Science"
author: "[name]"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

* You must submit your mini-assignment individually by end-of-class or end-of-day. Your name must exist in your mini-assignment and the names of your collaborators.
* Mini-assignments are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on mini-assignments, and no opportunities for revisions and make-up.

# Instructions

* For the exercises, please provide your answers in full sentences.
* Write your answers by replacing the text "[Write your answer here]".
* Please read the problems fully and carefully before answering.
* When finished, knit this `.Rmd` to `.html`.
* Submit this `.Rmd` and the recently knitted `.html`.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(nycflights13)
library(nycflights23)

# bind the two data frames
flights <- nycflights13::flights %>% 
  rbind(nycflights23::flights)
```

# Collaborators

::: {style="color: blue;"}
[Write your answer here]
:::

# Exercises

## Problem 1

How does the median flight delay time vary throughout the day for each origin airport, broken down by year?

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 2

On which day of the week should you depart to minimize the chance of a delay? Does the best day vary depending on the origin airport and the year?

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 3

What are the top five destination airports with the lowest median delay times, considering the number of flights as a selection criterion? Additionally, which month and time of day offer the lowest chance of delay? Compare results across different years.

::: {style="color: blue;"}
[Write your answer here]
:::
