---
title: "Counting Words"
subtitle: "Fundamentals of Data Science"
author: "[name]"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

* You must submit your mini-assignment individually by end-of-class or end-of-day. Your name must exist in your mini-assignment and the names of your collaborators.
* Mini-assignments are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on mini-assignments, and no opportunities for revisions and make-up.

# Instructions

* For the exercises, please provide your answers in full sentences.
* Write your answers by replacing the text "[Write your answer here]".
* Please read the problems fully and carefully before answering.
* When finished, knit this `.Rmd` to `.html`.
* Submit this `.Rmd` and the recently knitted `.html`.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)

# text related packages
library(tidytext)
```

# Collaborators

::: {style="color: blue;"}
[Write your answer here]
:::

# Exercises

## Problem 1

Below is an R code chunk that load the data sets `avatar-the-last-airbender.csv` and the `starwars.csv` files. The file [`avatar-the-last-airbender.csv`](https://github.com/averyrobbins1/appa){target="_blank"} contains transcripts of all three seasons of the show "Avatar: Last Airbender". The file [`starwars.csv`](https://github.com/kamran786/Star-Wars-Movie-Scripts-Text-Analysis){target="_blank"} contains the transcripts of the 6 Star Wars movies. The code below also loads the stopwords lexicon.

```{r message=FALSE, warning=FALSE}
# load the stopwords lexicon
swd_list <- tidytext::stop_words %>% 
  filter(lexicon == "SMART") %>%
  select(word)
swd_list <- swd_list$word

# load the data sets
atla <- read_csv("avatar-the-last-airbender.csv")
sw <- read_csv("starwars.csv")
```

a. Examine the two data sets and identify which is structured, unstructured, or semi-structured. List and identify the types of variables for each data set.

::: {style="color: blue;"}
[Write your answer here]
:::

b. Using the Starwars data and the `tidytext` package to tokenize the text data, count the words in the text variable for each season/book/movie, filter out the stopwords, and visualize the top 15 words using a barplot. Below, is an example code for you to get started.

```{r}
### [BEGIN] - CHANGE DATA SET HERE
df <- atla
### [END] - CHANGE DATA SET HERE

### [BEGIN] - WORK ON YOUR DATA WRANGLING HERE
# tokenize text and count words
df_counts <- df %>% 
  # group by season/book/movie
  group_by(book) %>%
  # automatically tokenizes the text and puts them into the word variable
  unnest_tokens(word, full_text) %>% 
  # use the count function to get the frequency of each word
  count(word) %>%
  # filter out the stopwords using the stopwords list defined earlier
  filter(!word %in% swd_list) %>%
  # count the frequency of each word in each group
  arrange(desc(n))
### [END] - WORK ON YOUR DATA WRANGLING HERE
```

::: {style="color: blue;"}
[Write your answer here]
:::
