---
title: "Transforming Tables for Visualizations"
subtitle: "Fundamentals of Data Science"
author: "MTH-391A | Spring 2025 | University of Portland"
date: "February 12, 2025"
output:
  slidy_presentation:
    font_adjustment: +5
    footer: "| MTH-391A Spring 2025 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
bibliography: ../../references.bib
csl: ../../mbe.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(gridExtra)
library(ggmosaic)
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- **Learn how to combine the `tidyverse` verbs and `ggplot2` geometries to visualize numerical and categorical data effectively**
- **Activity: Explore Different Visualization Techniques**
::::

:::: {.column width=15%}
::::

## Previously... (1/2)

**Chaining `dplyr` Verbs Using `|>`**

**Load Packages**

```{r eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```

**Define Data Frame as a Tibble**

```{r echo=TRUE, eval=TRUE}
iris_tibble <- tibble(iris)
```

**Advanced Example:** The goal of this example is to transform the `iris` dataset by computing the ratio of `Petal.Length` to `Sepal.Length` for observations belonging to the "setosa" species.

```{r echo=TRUE, eval=FALSE}
iris_tibble |>  
  # rule 1: choose only the "setosa" species
  filter(Species == "setosa") |>  
  # rule 2: pick the columns Sepal.Length and Petal.Length
  select(Sepal.Length,Petal.Length) |>  
  # rule 3: create a new column called length_ratio
  mutate(length_ratio = Petal.Length/Sepal.Length)
```

## Previously... (2/2)

**Visualizing Numerical Data using `ggplot2`**

**Example:** Plotting lengths by species in the `iris` data set.

```{r echo=TRUE, eval=TRUE, fig.align='center', out.width='100%'}
# establish data and variables
ggplot(iris_tibble, 
       aes(x = Sepal.Length, 
           y = Petal.Length,
           color = Species)) +
  # draw scatter plot
  geom_point() + 
  # add theme layer
  theme_grey()
```
::::

## Case Study 1: Titanic Data Set

The the **Titanic Data Set** is a popular dataset in data science. It contains information about the passengers aboard the Titanic and is often used for survival prediction.

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
# The titanic.csv data is from https://www.kaggle.com/datasets/aliaamiri/titanic-passengers-and-crew-complete-list
titanic <- read_csv("titanic.csv") |> 
  mutate(age = round(age,0),
         # recode the "crew" class of the class categorical variable using the `case_when` function
         class = case_when(
           class == "deck crew" ~ "crew",
           class == "engineering crew" ~ "crew",
           class == "restaurant staff" ~ "crew",
           class == "victualling crew" ~ "crew",
           TRUE ~ class
         ))
glimpse(titanic)
```

## Presenting Tables

To present tables effectively, we can use `kable` from the `knitr` package.

**Load Package**

```{r echo=TRUE, eval=TRUE}
library(knitr)
```

**Example:** The contingency table below shows the distribution of survival and different classes of passengers on the Titanic.

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
titanic_surv_class <- titanic |> select(survived,class) |> 
  group_by(survived,class) |> 
  summarise(total = n(),
            .groups = 'drop')
kable(xtabs(total ~ survived + class, titanic_surv_class))
```

## Bar Plots

A **bar plot** is a common way to display a single categorical variable. A bar plot where proportions instead of frequencies are shown is called a **relative frequency bar plot**.

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='80%', fig.width=6,fig.height=3}
p1 <- ggplot(titanic,aes(x=survived)) + 
  geom_bar() + 
  ggtitle("Frequencies") + 
  theme_minimal()
p2 <- ggplot(titanic,aes(x=survived)) + 
  geom_bar(aes(y=..count../sum(..count..))) + 
  ylab("proportion") + 
  ggtitle("Relative Frequencies") + 
  theme_minimal()
grid.arrange(p1, p2, ncol=2)
```

## Bar Plots with Two Variables

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='100%', fig.width=12,fig.height=4}
p1 <- ggplot(titanic,aes(x=class,fill=survived)) + 
  geom_bar() + 
  scale_fill_manual(values=c("yes" = "blue", "no" = "red")) + 
  ggtitle("Stacked") + 
  theme_minimal()
p2 <- ggplot(titanic,aes(x=class,fill=survived)) + 
  geom_bar(position="dodge") + 
  scale_fill_manual(values=c("yes" = "blue", "no" = "red")) + 
  ggtitle("Side-by-Side") + 
  theme_minimal()
p3 <- ggplot(titanic,aes(x=class,fill=survived)) + 
  geom_bar(position="fill") + 
  scale_fill_manual(values=c("yes" = "blue", "no" = "red")) + 
  ggtitle("Standardized") + 
  theme_minimal()
grid.arrange(p1, p2, p3, ncol=3)
```

::: {style="color: red;"}
$\star$ **Key Idea:** Each visualization provides a different perspective: Stacked helps compare total passengers across classes. Side-by-Side helps compare survival counts directly. Standardized helps compare survival rates across classes.
::::

## Mosaic Plots

A **mosaic plot** is a graphical representation of categorical data, displaying proportions and relationships between multiple categorical variables using a tiled area chart.

:::: {.column width=49%}
```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='100%', fig.width=4,fig.height=4}
ggplot(titanic) + 
  geom_mosaic(aes(x=product(class,survived), fill = survived)) + 
  scale_fill_brewer(palette = 1) +
  theme_mosaic() + 
  theme(legend.position = "none")
```
::::

:::: {.column width=50%}
**Features:** 

* Each tile represents a combination of categories from two or more categorical variables.
* The area of each tile is proportional to the frequency or probability of that combination.
* Provides an intuitive way to analyze relationships, such as independence or associations between categorical variables.
::::

## Pie Charts

A **pie chart** is a circular statistical graphic divided into slices to illustrate numerical proportions of a whole.

:::: {.column width=50%}
**Features:** 

* Each slice represents a category’s proportion relative to the whole dataset.
* The total sum of all slices equals 100%.
* Useful for showing part-to-whole relationships.
::::

:::: {.column width=49%}
```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='100%', fig.width=5,fig.height=5}
p1 <- ggplot(titanic |> filter(class=="1st"), aes(x="", y=survived, fill=survived)) +
  geom_bar(stat="identity", width=1) +
  scale_fill_brewer(palette = 1) +
  coord_polar("y", start=0) + 
  ggtitle("1st Class") + 
  theme_void()
p2 <- ggplot(titanic |> filter(class=="2nd"), aes(x="", y=survived, fill=survived)) +
  geom_bar(stat="identity", width=1) +
  scale_fill_brewer(palette = 1) +
  coord_polar("y", start=0) + 
  ggtitle("2nd Class") + 
  theme_void()
p3 <- ggplot(titanic |> filter(class=="3rd"), aes(x="", y=survived, fill=survived)) +
  geom_bar(stat="identity", width=1) +
  scale_fill_brewer(palette = 1) +
  coord_polar("y", start=0) + 
  ggtitle("3rd Class") + 
  theme_void()
p4 <- ggplot(titanic |> filter(class=="crew"), aes(x="", y=survived, fill=survived)) +
  geom_bar(stat="identity", width=1) +
  scale_fill_brewer(palette = 1) +
  coord_polar("y", start=0) + 
  ggtitle("Crew") + 
  theme_void()
grid.arrange(p1, p2, p3, p4, nrow=2, ncol=2)
```
::::

## Comparing Categorical Data to Numerical Data

**Example:** Age distribution among class passengers and survival. 

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='80%', fig.width=8,fig.height=4}
ggplot(titanic,aes(y=class,x=age, fill=survived)) + 
  geom_boxplot() + 
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()
```

## Adding More Variables

**Example:** Age distribution among class passengers and survival. 

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='80%', fig.width=8,fig.height=4}
ggplot(titanic,aes(y=fare,x=age, color=survived, pch = gender)) + 
  geom_point() + 
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()
```

## Activity: Visualize Multiple Variables

1. Log-in to Posit Cloud and open the R Studio assignment *MA8: Visualize Multiple Variables*.
2. Make sure you are in the current working directory. Rename the `.Rmd` file by replacing `[name]` with your name using the format `[First name][Last initial]`. Then, open the `.Rmd` file.
3. Change the author in the YAML header.
4. Read the provided instructions.
5. Answer all exercise problems on the designated sections.
