---
title: "Visualize Multiple Variables"
subtitle: "Fundamentals of Data Science"
author: "[First name][Last initial]"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

* You must submit your worksheet individually by end-of-class or end-of-day. Your name must exist in your worksheet and the names of your collaborators.
* Worksheets are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on worksheets, and no opportunities for revisions and make-up.

# Instructions

* For the exercises, please provide your answers in full sentences.
* Write your answers by replacing the text "[Write your answer here]".
* Please read the problems fully and carefully before answering.
* When finished, knit this `.Rmd` to `.html`.
* Submit this `.Rmd` and the recently knitted `.html`.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(gridExtra)
library(ggmosaic)
```

# Collaborators

::: {style="color: blue;"}
[Write your answer here]
:::

# Exercises

## Problem 1

Load the `titanic.csv` data set using the `read_csv()` function.

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 2

Examine the Titanic data set and create a contingency table that represents the number of survived and non survived males and females.

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 3

Generate a plot that shows the proportion of survived males and females compared to not survived male and females.

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 4

Generate a plot that shows the `fare` variable comparing different genders and survival.

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 5

Generate a plot that shows the `age` variable comparing different classes and survival.

::: {style="color: blue;"}
[Write your answer here]
:::
