---
title: "Data Wrangling on Multiple Tables"
subtitle: "Fundamentals of Data Science"
author: "MTH-391A | Spring 2025 | University of Portland"
date: "February 7, 2025"
output:
  slidy_presentation:
    font_adjustment: +5
    footer: "| MTH-391A Spring 2025 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
bibliography: ../../references.bib
csl: ../../mbe.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- **Know how to load external data files in R**
- **Develop an understanding of combining data frames**
- **Activity: Combine Datasets**
::::

:::: {.column width=15%}
::::

## Loading an External Data Set

**Load packages**

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
```

**Importing a  (CSV) `.csv` file**

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
birthdays <- read_csv("birthdays.csv",col_names=TRUE)
```

::: {style="color: red;"}
$\star$ Note that you need to have the `birthdays.csv` file in your working directory.
:::

**Importing a (Excel) `.xlxs` file**

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
us_capitals <- read_excel("usa-state-capitals.xlsx", sheet = "owssvr")
```

::: {style="color: red;"}
$\star$ Note that you need to have the `usa-state-capitals.xlsx` file in your working directory.
:::

## Other File Types

| *Verb* | *Function* |
|:---|:------|
| `read_tsv()` | Reads tab-delimited files |
| `read_delim()` | Reads in files with any delimiter, attempting to automatically guess the delimiter if you don’t specify it |

::: {style="color: red;"}
$\star$ Note that data sets are commonly saved as CSV and Excel files but there are other filetypes (such as shapefiles) that require special functions, which we will discuss soon.
:::

## View the Imported Data Sets

:::: {.column width=49%}
```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
birthdays
```
::::

:::: {.column width=50%}
```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
us_capitals
```
::::

::: {style="color: red;"}
$\star$ It looks like the `state` variable in the `birthdays` data frame have categories matching the `ST` variable in the `us_capital` data frame.
:::

## Using `inner_join()`

**Joining `birthdays` and `us_capital` by matching the state abbreviations**

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
states_joined <- birthdays |>
  inner_join(us_capitals, by = c("state" = "ST"))
states_joined
```

::: {style="color: red;"}
$\star$ The result contains only those rows that have matches in both tables.
:::

## Using `left_join()`

**Joining `us_capitals` into `birthdays` by matching the state abbreviations**

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
states_leftjoined <- birthdays |>
  left_join(us_capitals, by = c("state" = "ST"))
states_leftjoined
```

::: {style="color: red;"}
$\star$ The rows of the first table are always returned, regardless of whether there is a match in the second table.
:::

## Using `right_join()`

**Joining `birthdays` into `us_capitals` by matching the state abbreviations**

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
states_rightjoined <- birthdays |>
  right_join(us_capitals, by = c("state" = "ST"))
states_rightjoined
```

::: {style="color: red;"}
$\star$ Notice here the there result is the same the using the `inner_join()` because the `us_capital` only includes all 50 states while the `birthdays` include all 50 states plus DC.
:::

## Activity: Combine Datasets

1. Log-in to Posit Cloud and open the R Studio assignment *MA6: Combine Datasets*.
2. Make sure you are in the current working directory. Rename the `.Rmd` file by replacing `[name]` with your name using the format `[First name][Last initial]`. Then, open the `.Rmd` file.
3. Change the author in the YAML header.
4. Read the provided instructions.
5. Answer all exercise problems on the designated sections.
