---
title: "Categorizing Word Sentiments"
subtitle: "Fundamentals of Data Science"
author: "[name]"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

* You must submit your mini-assignment individually by end-of-class or end-of-day. Your name must exist in your mini-assignment and the names of your collaborators.
* Mini-assignments are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on mini-assignments, and no opportunities for revisions and make-up.

# Instructions

* For the exercises, please provide your answers in full sentences.
* Write your answers by replacing the text "[Write your answer here]".
* Please read the problems fully and carefully before answering.
* When finished, knit this `.Rmd` to `.html`.
* Submit this `.Rmd` and the recently knitted `.html`.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)

# text related packages
library(tidytext)
library(wordcloud)
```

# Collaborators

::: {style="color: blue;"}
[Write your answer here]
:::

# Exercises

## Problem 1

Below is an R code chunk that load the data set `avatar-the-last-airbender.csv` file. The file [`avatar-the-last-airbender.csv`](https://github.com/averyrobbins1/appa){target="_blank"} contains transcripts of all three seasons of the show "Avatar: Last Airbender".

```{r message=FALSE, warning=FALSE}
# load the stopwords lexicon
swd_list <- tidytext::stop_words %>% 
  filter(lexicon == "SMART") %>%
  select(word)
swd_list <- swd_list$word

# sentiment lexicon
emotions_2 <- tidytext::sentiments # positive/negative words
emotions_10 <- tidytext::get_sentiments(lexicon=c("nrc")) # 10 emotions

# load the data
atla <- read_csv("avatar-the-last-airbender.csv")
```

Below is an R code that takes a subset of the `atla` data frame where it chooses two characters and tokenizes the words in the `character_words` variable while grouping them by `book` and `character` variables. Your task is to modify the code as listed below:
    
a. Include two other characters, and remove the stopwords.
b. Choose the positive and negative words in the `emotions_2` data frame and use it to subset the previous data frame. Hint: Use the `left_join()` function.
c. Create a standardized stacked barplot that compares the proportion of positive and negative words for each character in each book. Write a paragraph describing your observations.
d. Create a word cloud for each character and color each word according to its corresponding sentiment.

```{r}
atla_sub <- atla %>% 
  # pick characters
  filter(character %in% c("Aang","Zuko")) %>%
  # group by book and character
  group_by(book,character) %>%
  # tokenize the character's words and put it into the word variable
  unnest_tokens(word,character_words) %>%
  # count the words
  count(word)
```

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 2

Repeat the same process as Problem (1) but use the `emotions_10` lexicon.

a. How many emotions are included in the lexicon? What are they?
b. Create a standardized stacked barplot that compares the word proportions of your chosen 3 emotions for each chosen character in each book. Write a paragraph describing your observations.
c. Create a word cloud for each character and color each word according to its corresponding sentiment.

::: {style="color: blue;"}
[Write your answer here]
:::
